---
title: "Example Analysis"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Example Analysis}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Load the survival package and other necessary packages for analysis
```{r setup}
library(survival)
library(ggsurvfit)
library(tidyverse)
library(tidytuesdayR)
```

# Read the data
```{r}
create_dir <- function(){
  setwd(dirname(getwd()))
  data_dir <- "data"
  
  if(!file.exists(data_dir)){
    dir.create(data_dir)
  }
}

save_data <- function(){
  data_path <- file.path("data", "tuesdata.rds")
  
  if(!file.exists(data_path)){
    data <- readr::read_csv("https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2019/2019-10-08/ipf_lifts.csv")
    write.table(data, data_path)
  }
}

load_data <- function(){
  create_dir()
  save_data()
  data_path <- file.path("data", "tuesdata.rds")
  data <- read.table(data_path)
  return(data)
}

ipf_lifts = load_data()
```

The dataset I chose to use from TidyTuesday is the powerlifting dataset from 2019. The dataset includes information about powerlifters at specific competitions. It includes the competitor's name, sex, age, age class, event, equipment used, date of competition, bodyweight in kg, and the best of three lifts in each of the events (squat, bench press, and deadlift) in kg.

The question I aim to answer is whether age and bodyweight class impact the amount of time since a competitor's first recorded competition it takes a male competitor to reach the "overall international elite" status, as defined by the United States Powerlifting Association.

The data came from the TidyTuesday podcast. The original data as well as the data dictionary can be found at https://github.com/rfordatascience/tidytuesday/tree/master/data/2019/2019-10-08

# Analysis
```{r}
#Converting kgs to lbs, only selecting necessary variables, only including men who competed in all three events
ipf_lifts_lbs = ipf_lifts %>%
  mutate(bodyweight_lb = 2.205*bodyweight_kg,
         squat_lb = 2.205*best3squat_kg,
         bench_lb = 2.205*best3bench_kg,
         deadlift_lb = 2.205*best3deadlift_kg) %>%
  select(name, sex, event, equipment, age, age_class, date, bodyweight_lb, squat_lb, bench_lb, deadlift_lb) %>%
  filter(sex == "M", event == "SBD")


#Creating a binary variable that equals 1 if the lifter was categorized as "elite" during this event, and 0 otherwise.

#The criteria for "international elite" classification for men was found at https://www.lift.net/2013/05/09/classification-standards-for-raw-elite-uspa/

ipf_lifts_event = ipf_lifts_lbs %>%
  mutate(elite = ifelse(sex == "M" & bodyweight_lb <114 & squat_lb+bench_lb+deadlift_lb >=1044, 1,
                        ifelse(sex == "M" & bodyweight_lb >=114 & bodyweight_lb <123 & squat_lb+bench_lb+deadlift_lb >=1135, 1,
                               ifelse(sex == "M" & bodyweight_lb >=123 & bodyweight_lb <132 & squat_lb+bench_lb+deadlift_lb >=1222, 1,
                                      ifelse(sex == "M" & bodyweight_lb >=132 & bodyweight_lb <148 & squat_lb+bench_lb+deadlift_lb >=1368, 1,
                                             ifelse(sex == "M" & bodyweight_lb >=148 & bodyweight_lb <165 & squat_lb+bench_lb+deadlift_lb >=1498, 1,
                                                    ifelse(sex == "M" & bodyweight_lb >=165 & bodyweight_lb <181 & squat_lb+bench_lb+deadlift_lb >=1611, 1,
                                                           ifelse(sex == "M" & bodyweight_lb >=181 & bodyweight_lb <198 & squat_lb+bench_lb+deadlift_lb >=1698, 1,
                                                                  ifelse(sex == "M" & bodyweight_lb >=198 & bodyweight_lb <220 & squat_lb+bench_lb+deadlift_lb >=1790, 1,
                                                                          ifelse(sex == "M" & bodyweight_lb >=220 & bodyweight_lb <242 & squat_lb+bench_lb+deadlift_lb >=1854, 1,
                                                                                ifelse(sex == "M" & bodyweight_lb >=242 & bodyweight_lb <275 & squat_lb+bench_lb+deadlift_lb >=1909, 1,
                                                                                       ifelse(sex == "M" & bodyweight_lb >=275 & bodyweight_lb <308 & squat_lb+bench_lb+deadlift_lb >=1995, 1,
                                                                                              ifelse(sex == "M" & bodyweight_lb >=308 & squat_lb+bench_lb+deadlift_lb >=2055, 1, 0)))))))))))))

table(ipf_lifts_event$elite)

ggplot(data = ipf_lifts_time, aes(x = equipment, y = squat_lb)) + geom_violin() + labs(y = "Best squat (lbs)", title = "Violin plots of competitors' squat performances", subtitle = "In pounds, by equipment used", caption = "Violin plot of competitors' best squat lift in three tries. Separated by equipment type.")

ggplot(data = ipf_lifts_time, aes(x = bench_lb, group = equipment, fill = equipment)) + geom_boxplot() + labs(x = "Best bench (lbs)", title = "Boxplots of competitors' bench performances", subtitle = "In pounds, by equipment used", caption = "Boxplots of competitors' best bench lift in three tries. Separated by equipment type.")

g = ggplot(data = ipf_lifts_time, aes(x = bodyweight_lb, y = deadlift_lb)) + geom_point()
g + facet_wrap(vars(equipment)) + labs(x = "Bodyweight", y = "Best deadlift (lbs)", title = "Scatterplots of best deadlifts vs bodyweight", subtitle = "In lbs, separated by equipment type", caption = "Scatterplots of best deadlifts vs bodyweight in lbs, separated by equipment type")

#Out of the men that competed in all three events (squat, bench press, and deadlift), the "international elite" classification was reached 7971 times.


ipf_lifts_time$elite = compact(ipf_lifts_time$elite)
some(ipf_lifts_time$elite, is.na)
every(ipf_lifts_time$elite, is.na)
#So some of the events are NA, but not all of them.


#Now creating a "time" variable that measures the number of days since the first recorded competition for each player
ipf_lifts_time = ipf_lifts_event %>%
  group_by(name) %>%
  mutate(time = as.numeric(date)-min(as.numeric(date)))

ipf_lifts_time %>%
  group_by(name) %>%
  summarize(max(time))

ggsurvfit(survfit(Surv(time, elite) ~ 1, data = ipf_lifts_time))

#using age class
fct_relevel(ipf_lifts_time$age_class, "18-19", after = 2)

summary(coxph(Surv(time, elite) ~ age_class + bodyweight_lb, data = ipf_lifts_time))

#Using continuous age
summary(coxph(Surv(time, elite) ~ age + bodyweight_lb, data = ipf_lifts_time))
```

Overall, the Kaplan-Meier plot shows that as time since first competition increases, the probability of getting to "international elite" status decreases. Specifically, about 3000 days after the first competition, the probability of reaching "international elite" status drops to 0.5.

To see whether age and bodyweight affect the time until international elite status is reached, I ran a Cox proportional hazards model using time to international elite status as the outcome and age and bodyweight as predictors. I found that age significantly increases the time until international elite status is reached, adjusting for bodyweight.


Functions used
dplyr::select
dplyr::mutate
dplyr::filter
dplyr::group_by
dplyr::summarize
forcats::fct_relevel

purrr::compact
purrr::some
purrr::every

ggplot2::geom_violin
ggplot2::geom_boxplot
ggplot2::geom_point
ggplot2::facet_wrap

survival::survfit
survival::coxph
